\clearpage{}


\section*{{\large{}Reading and Resources}}


\subsection*{Python:}
\begin{itemize}
\item ``A Hands-On Introduction to Using Python in the Atmospheric and
Oceanic Sciences'' by Johnny Wei-Bing Lin. Download for free or buy
cheaply from \url{http://www.johnny-lin.com/pyintro} .
\item ``Think Python. How to Think Like a Computer Scientist'' by Allen
B. Downey . Download from free from \url{http://www.greenteapress.com/thinkpython/thinkpython.html}
\item ``Numerical Methods in Engineering with Python'' by Jaan Kiusalaas 
\item \url{www.southampton.ac.uk/sesg3020/textbook/SciPyIntro.pdf} 
\item \url{http://www.python.org} 
\item \url{http://docs.python.org/tutorial/} 
\item \url{http://matplotlib.sourceforge.net/users/index.html} 
\item \url{http://matplotlib.sourceforge.net/users/pyplottutorial.html} 
\item \url{http://software-carpentry.org/} 
\item \url{http://www.diveintopython.net/} 
\item \url{http://learnpythonthehardway.org/} 
\item \url{http://www.ibiblio.org/g2swap/byteofpython/read/} 
\item ``Making Use of Python'' by Rashi Gupta 
\item ``Python Essential Reference'' by David M. Beazley (Addison Wesley) 
\item ``Learning Python'' Mark Lutz (O'Reilly Media) 
\end{itemize}
\clearpage{}


\subsection*{Version Control}
\begin{itemize}
\item \url{https://github.com/}
\item \url{https://swcarpentry.github.io/git-novice/}
\item \url{http://product.hubspot.com/blog/git-and-github-tutorial-for-beginners}
\end{itemize}

\subsection*{Numerical Methods: }
\begin{itemize}
\item Wikipedia \url{http://en.wikipedia.org} 
\item Durran, D. R. Numerical methods for fluid dynamics (Springer). 
\item LeVeque, R. Numerical Methods for Conservation Laws (Springer)
\item Ortega, J.M. and Poole, W.G. An Introduction to Numerical Methods
for Differential Equations. 1981 (Pitman Publishing Inc)
\item Ferziger, J. H. and Peric, M. Computational Methods for Fluid Dynamics
(Springer). 
\item Numerical Recipes: The Art of Scientific Computing, Third Edition
(2007), (Cambridge University Press). \url{http://www.nr.com/} 
\item Lloyd N. Trefethen. Finite Difference and Spectral Methods for Ordinary
and Partial Differential Equations. \url{https://people.maths.ox.ac.uk/trefethen/pdetext.html}
\end{itemize}

\subsection*{Website:}

The lecturer's version of the notes be downloaded from\\
 \url{http://www.met.reading.ac.uk/~sws02hs/teaching/WellerSSnotes_2_lec.pdf}


\subsection*{Highly Recommended}

{\color{red} Install Python on your own computer. It is free. See,
eg

\url{http://code.google.com/p/pythonxy/} }
