
\chapter{Software Architecture}

There are numerous very bad programming practices in science. We will
cover a few ways in which things could be done better. 

\input{goodCode.tex}


\section{Testing\label{sec:testing}}

Code should be tested in situations where you know exactly what the
output should. This is a crucial part of writing bug-free code. Every
code and every function within a code should be tested. For a function,
you should think of a variety of different inputs for which you know
the output and write a tests that calls the function with those inputs
and checks that the outputs are as expected. There may also be situations
where a function is expected to give an error message. The tests should
include input that you know should give an error message.

There are python modules to help with testing such as \url{unittest}
and \url{nosetest}. Tools like \url{coverage} can show you how much
of your code you have tested. There is good integration with some
integrated development environments (IDEs), such as \url{PyCharm},
which helps with most aspects of writing, testing, debugging, and
running code.

\clearpage{}


\section{Debugging}

\stepwise{

IDEs (Interactive Development Environments, such as PyCharm) usually
have useful interactive debugging tools which allow you to set breakpoints
in the code and see, for example, what all variable values are. There
are also python libraries such as \url{pdb} which stop the code at
the line \url{pdb.set_trace()} and then allow you to step through
and see what all variable values are. You should also be able to debug
by hand without a dedicated tool, using lots of print statements.
Here are some tips:
\begin{itemize}
\item \step{You should try to re-produce the problem with the simplest
code and with arrays set as small as possible so that the code runs
(and crashes) quickly and so that you can re-produce the calculations
by hand in order to check what the code is doing.}\step{
\item You should have some idea of where things go wrong. Start by putting
in print statements at a point in the code shortly before everything
went wrong. Print out variables that you think might be related to
the problem and check that their values are as expected. }\step{
\item Put in print statements further up where you think that everything
was working fine, printing out relevant variables.}\step{
\item Using more print statements, find where in between things go wrong.
}\step{
\item When you have found and corrected the bug, delete all the de-bugging
code rather than commenting it out. You do not want your code littered
with commented out code. You might like to save all of the debug statements
in a branch of a version control repository so that you can refer
to it later. }\step{
\item Think how this bug happened and how you can better test your code
in future to avoid bugs like that happening again.}
\end{itemize}
}\clearpage{}


\section{Version Control using git\label{sec:git}}

Very quickly after starting to write code, you will find that you
have modified the code and it doesn't do what you want any more, you
have multiple versions and you can't keep track, two people are working
on the same code and getting in a muddle, you cannot reproduce the
results that you got yesterday. You need to be using version control
software like \url{git} and a repository hosting service like \url{https://github.com/},
which also provides archiving and backup.

\pause There are numerous online resources for how to use \url{git}
and \url{github}. This will therefore be a tiny introduction. A more
comprehensive introduction is at, for example:\\
\url{https://swcarpentry.github.io/git-novice/}\\
and\\
\url{http://product.hubspot.com/blog/git-and-github-tutorial-for-beginners}.
You should really also know about collaborating, branching and merging
in \url{git} but this little introduction will cover setting up a
repository, adding to it, pushing and pulling from a remote repository
and recovering old versions. This introduction draws heavily from
the Software Carpentry introduction. \pause 


\subsection{Setting up \protect\url{git}}

These are the commands that I used to set up \url{git} for my login
on the machines at Cambridge:

\begin{lstlisting}
$ git config --global user.name "Hilary Weller"
$ git config --global user.email "h.weller@reading.ac.uk"
$ git config --global color.ui "auto"
$ git config --global core.editor "gedit --wait --new-window"
\end{lstlisting}


\clearpage{}


\subsection{Setting up a Local Repository}

Next set up a local repository for my \url{helloWorld.py} code:

\begin{lstlisting}
# Initialise the local directory as a git repository
$ git init
# Add files to the repo, check and commit them with a commit message
$ git add helloWorld.py
# Now the file is staged for commit. This can be checked and committed:
$ git status
$ git commit -m "First commit"
\end{lstlisting}
\pause 


\subsection{Sign up for a \protect\url{github} account and create a Remote Repository}
\begin{itemize}
\item Sign up for a free account allowing unlimited public code at \url{https://github.com/}
\item Create a new repository at \url{https://github.com/new} or by following
the links. I created a repository called \url{helloWorld}. My new
\url{github} user name is \url{studentSpencer0}. 
\end{itemize}
\pause 


\subsection{Link the local and remote repositories}

Copy the url of your new repository. Mine is \url{https://github.com/studentSpencer0/helloWorld}.
Then, on the command line:
\begin{lstlisting}
$ git remote add origin https://github.com/studentSpencer0/helloWorld
$ git remote -v     # (to verify the new remove URL)
\end{lstlisting}
\pause 


\subsection{Push the changes in your local repository to GitHub.}

If you initialised the remote repository with a README pull and merge
before push
\begin{lstlisting}
$ git pull origin master
\end{lstlisting}
After closing the window that asks you to write a merge message you
can now do:

\begin{lstlisting}
$ git push origin master
\end{lstlisting}
 Now have a look at the url for your repository and see your new code
there.

\clearpage{}


\subsection{Some more configuration}

To avoid always having to specify
\begin{lstlisting}
origin master
\end{lstlisting}
when pushing and pulling, you can set some more defaults
\begin{lstlisting}
$ git config --global pull.default matching
$ git branch --set-upstream-to=origin/master master
\end{lstlisting}
Now to push and pull you only need:
\begin{lstlisting}
$ git pull
$ git push
\end{lstlisting}
You can avoid entering your \url{github} username and password every
time you push by typing:
\begin{lstlisting}
git config credential.helper store
\end{lstlisting}
\pause 


\subsection{Adding all suitable files}

We may have a directory structure full of files that we want to add
to the repository. However it is important not to add large or binary
files to a git repository. Therefore we need a \url{.gitignore} file
to tell git not to add files with well known extensions that imply
that they are binary files. You can copy my \url{.gitignore} file:

\begin{lstlisting}[basicstyle=\scriptsize]
$ wget https://raw.githubusercontent.com/studentSpencer0/helloWorld/master/.gitignore
\end{lstlisting}

You can now add all files in the directory structure to the repository
without the risk of adding large or binary files
\begin{lstlisting}
$ git add .
\end{lstlisting}
\clearpage{}


\subsection{Retrieving Old Versions}
\begin{itemize}
\item In order to see a list of commit ids and messages:
\begin{lstlisting}
$ git log
\end{lstlisting}
\pause 
\item If you make some changes to a file (say \url{helloWorld.py}), you
can see the differences between what is on disk and the version of
the previous commit:
\begin{lstlisting}
$ git diff helloWorld.py
# or if you have already added helloWorld.py but not committed:
$ git diff HEAD helloWorld.py
# or if you want to compare helloWorld.py to the version from 
# three commits ago:
$ git diff HEAD~3 helloWorld.py
\end{lstlisting}
\pause 
\item You can also show the difference between a file on disk and a previous
commit using the commit id:
\begin{lstlisting}
$ git diff cb27fc3e526dbdcd49b70b3f08a520ae271f8be5 helloWorld.py
\end{lstlisting}
\pause 
\item To show what changes were made at an older commit rather than differences
between working directory and a commit:
\begin{lstlisting}
$ git show HEAD~3 helloWorld.py
\end{lstlisting}
\pause 
\item If you need to discard the changes that you have made and go back
to the most recent commit:
\begin{lstlisting}
$ git checkout -- helloWorld.py
\end{lstlisting}
or to go back to a particular commit (just using first few letters
of the commit id):
\begin{lstlisting}
$ git checkout f22b25e helloWorld.py
\end{lstlisting}
\end{itemize}

